Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-14 csf-2017/paper.tex
udpated f1 and f2 paper to have the programs side by side
Add a discussion of stack clearing and fix a missing requirement in the stack discipline: stack needs to be cleared before returning...

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item In Section~\ref{sec:capab-mach-with}, we formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack in Section~\ref{sec:stack-and-return-pointer}. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., (potentially)
  cross-component function pointers) and can be efficient assuming
  only
  one additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed
  logical relation with recursively defined Kripke worlds. It is
  inspired by one used in \cite{Devriese:2016ObjCap} for
  capability safety in a JavaScript-like lambda calculus, but has two
  novel technical characteristics. The natural CPS style of assembly
  code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal
  closure that has been previously used for lambda calculi. Perhaps
  more surprisingly, we show that the special nature of local
  capabilities can be elegantly expressed using a variant of an idea
  that has been used by Dreyer et. al.~\cite{Dreyer:jfp12} to express
  well-bracketedness guarantees in a lambda calculus: Kripke worlds
  with both a public and a private future world relation. Interestingly,
  the two future world relations generally work very similarly to
  Dreyer et al.'s, except that they turn up in quite a different place
  in the logical relation. The logical relation and theorem is presented in Section~\ref{sec:logical-relation}.
\item We demonstrate the power of our results in Section~\ref{sec:examples} by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability safety and our stack management discipline.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

The $\load{}{}$ instruction loads the contents of the requested memory location
into the appropriate register, but only if the capability has appropriate
authority (i.e.\ read permission and an appropriate range). The $\restricttwo{}{}$
instruction will update a capability's permissions and global/local status in
the register file, but requires that the new permissions are weaker than the
original and does not turn local capabilities into global ones. The $\geta{}{}$
instruction queries the current address of a capability and stores it in another
register.

@@

The final semantics we discuss is that of $\store{}{}$. As expected, it updates
the memory to the requested value if the capability has write authority for the
requested location. However, the instruction is complicated by the presence of
\emph{local capabilities}, modeled after the ones in
CHERI~\cite{Watson2015Cheri}. Basically, local capabilities are special in that
they can only be kept in registers, i.e.\ they cannot be stored to memory. This
means that local capabilities can be \emph{temporarily} given to an adversary,
for the duration of an invocation: if we take care to erase the capability from
the register file after control is passed back to us, we can be sure that they
will not have been able to store the capability. However, there is one exception
to the rule above: local capabilities can be stored to a part of memory for
which we have a capability with write-local authority (i.e.\ permission $\rwl$ or
$\rwlx$). This exception is intended to accomodate regions of memory that behave
like the C stack: stashing away the contents of registers on the stack can be
allowed, even if this includes local capabilities. As long as all capabilities
with write-local authority are themselves local and the stack is cleared after
control is passed back by the adversary, we will see that this does not break
the intended behavior of local capabilities.

@@

\textbf{Program startup}
When our code starts executing, we assume that the language runtime initializes
the memory as follows: a contiguous array of memory is reserved for the stack,
for which we receive a stack pointer in a register, $r_\stk$, that we reserve for this
purpose. We stress that the stack is not built-in, but merely an
abstraction we put on a piece of the memorey. This stack pointer must be local and it must carry $\rwlx$ permission. It is crucial for
our approach that the stack is the only part of memory for which there will ever
be an obtainable capabilities with $\rwlx$ or $\rwl$ capabilities. We expect the language
runtime to enforce this by not handing out such capabilities for other parts of
memory in any situation (i.e.\ not through malloc, not during loading or linking
etc.). Additionally, in our examples, we typically also assume some static parts
of memory where instructions are stored or static data. Another part of memory
(called the heap) is initially governed by malloc and at program startup, no
other code has capabilities for this memory. Malloc will hand out capabilities
for allocated regions as requested, but it will never hand out capabilities with
write-local capability (i.e.\ no $\rwlx$ or $\rwl$ permissions). For simplicity,
we assume that memory allocated through malloc cannot be freed.

\textbf{Returning to the adversary} Perhaps the simplest scenario is the case of
returning to the adversary after they invoked our code. In this case, we will
have received a return pointer from them, and we just need to jump to it as
usual. The only thing we need to take care of is to properly clear the registers
before we do (erasing any data we do not want to pass as (part of) the return
value) and similarly, erase any part of the stack that we have used (i.e.\
overwrite it with zeros). Note that we can use the stack as normal: we may push
values onto it and pop them back off (e.g. for register spilling etc.).

@@

A final required safety measure is that before invocation of the adversary, we
need to clear (i.e.\ overwrite with zeros) the part of the stack that we are not
using. Otherwise, that part of the stack may still contain data and capabilities
from previously executed code, including local capabilities since they are
allowed on the stack. The adversary could take advantage of this by using it to
store either stack pointers or return pointers from previous invocations and
play the same tricks as above. Clearing the entire unused part of the stack is
the most problematic aspect of our discipline in terms of efficiency, since it
may consist of a large part of memory. We believe that capability machines like
CHERI should consider to provide special support for this requirement, by
implementing an optimized instruction for clearing
a large block of memory. We have no experience in processor design, but we
imagine this might be optimized as a vector instruction or by having the
processor keep track of a range of memory for which loads should return zero
irrespective of the actual contents of the memory.

@@

It's important that neither of these capabilities are accessible to the
adversary. In other words, the return pointer provided to the adversary must be
a capability that they can jump to but not read from, i.e.\ an enter
capability. To make this work, we construct the activation record as depicted in
Figure~\ref{fig:activ-rec-struct}. The return pointer is an $\entry$ capability
with authority over the activation record containing the previous return and
stack pointer as well as a number of restore instructions. The return pointer's
current address points to those instructions, so that, upon invocation, they are
executed and can load the old stack pointer and program counter back into the
register file. Because the return pointer is given to the adversary as an enter
pointer, they cannot get hold of the contents of the data structure, but after
invocation, its permission is updated to $\exec$, so the activation record's
contents become available to the restore instructions.

The final question that remains is: where should we store this activation
record? The attentive reader may already see that there is only one possibility:
since the activation record contains the old stack pointer, which is necessarily
local, the activation record can only be constructed in a part of memory where
we have write-local access, i.e.\ on the stack. Note that this means we will be
placing and executing instructions on the stack, i.e.\ it will not just contain
code pointers and data. This means that our discipline should be combined with
protection against stack smashing attacks (i.e.\ buffer overflows on the stack
overwriting activation records' contents), but conveniently, the fine-grained
memory protection on capability machines is well-suited to easily provide such
protection.

@@

Our answer is that the adversary should provide it to us, much like we provide
them with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
capability and make sure its current address points just below its start address.
Without this check, an adversary could potentially get control over
our local stack frame during a callback by passing us a local $\rwx$
capability to a global data structure (i.e.\ there exists a global
capability for it) as the stack pointer and a global
return pointer. If our local state contains no local capabilities,
then following our calling convention, the callback would not fail and
the adversary could use a stored capability for the global data
structure to access our local state. 
To prevent this from happening, we need to make sure the stack
capability is local and there exists no global capability for that
memory segment. By making sure the capability has permission $\rwlx$,
we rely on the system wide assumption that there are only local
$\rwlx$ capabilities.

\textbf{Stack management discipline}
The precautions the needed to safely interact with an adversary have now been introduced and motivated. The stack discipline comprises the precautions. Summarized, they were:
\begin{description}[font=\normalfont\itshape]
\item[At program startup] a local $\rwlx$-capability (the stack pointer) resides in register $r_\stk$. The system does not hand out global write-local capabilities.
\item[Before returning to the adversary] clear the registers (with words we do not want to pass to the adversary) and erase any part of the stack we have used (any part of the stack that has been passed to an adversary must also be erased).
\item[Before invoking the adversary] push an activation record to the stack, create a local $\entry$-capability for the restore instructions of the activation record (the return pointer) and leave it in a register, leave a copy of the stack capability restricted to the unused part in $r_\stk$, erase the part of the stack we release control over, and clear the non-argument registers.
\item[Before invoking an adversary callback] make sure it is a global capability.
\item[When called by an adversary] make sure the alleged stack pointer has permission $\rwlx$.
\end{description}
\textbf{Reusable macro instructions} 
We define a number of reusable macros. Some of them capture the stack
management discipline and others make the examples in
Section~\ref{sec:examples} readable. The macros are:
\texttt{fetch $r$ $\var{name}$} fetches the capability related to
$\var{name}$ from the linking table and stores it in register
$r$. \texttt{push $r$}/\texttt{pop $r$} works as one would expect. All
macros that uses the stack assumes a stack pointer in register
$r_\stk$.  \texttt{prepstk $r$} is used when a callback is invoked by
the adversary and prepares the received stack pointer by checking that
it has permission $\rwlx$ and making it conform to our stack
discipline. \texttt{scall
  $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)} Jumps
to the capability in register $r$ in the manner described above. That
is, it pushes local state (the contents of registers
$\overline{r_{\var{priv}}}$), stack pointer and return pointer to the
stack, sets up the necessary on stack return code and creates a return
pointer, restricts the stack pointer, clears the unused part of the
stack, clears the necessary registers and jumps to $r$. Upon return,
the private state is restored. \texttt{mclear $r$} clears all the
memory the capability in register $r$ has authority
over. \texttt{rclear $\var{regSet}$} clears all the registers in
$\var{regSet}$. \texttt{reqglob $r$} checks whether the word in
register $r$ is a global capability. \texttt{crtcls
  $\overline{(x_i,r_i)}$ $r$} allocates a closure pair where the
capability in register $r$ is used to point to the program and a new
environment is allocated (using malloc) where the contents of
$\overline{r_i}$ is stored. In the program referred to by $r$, an
implicit fetch happens when an instruction refers to $x_i$.

@@

% \begin{figure}[htbp]
%   \centering
%   \begin{align*}
%   &\iota_{\start,\addrend}^\pwl : \Regions\\
%   &\iota_{\start,\addrend}^\pwl \defeq (\temp,1,=,=,H^\pwl_{\start,\addrend}) \\
%   \\
%   &H^\pwl : \Addrs^2 \fun \States \fun (\Wor \monwknefun \UPred{\HeapSegments})\\
%   &H^\pwl_{\start,\addrend}\; s \; \hat{W} \defeq  \quad\left\{\npair{\hs} \; \middle| \;
%     \begin{aligned}
%       & n = 0 \text{\footnotesize{ or }} (\dom(\hs) = [\start,\addrend] \text{\footnotesize{ and }} \\
%       &\forall \addr \in [\start,\addrend] \ldotp\\
%       & \;\; \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W})))
%     \end{aligned}
%         \right\}\\
%   &\iota_{\start,\addrend}^{\nwl} : \Regions \\
%   &\iota_{\start,\addrend}^{\nwl} \defeq (\temp,1,=,=,H^\nwl_{start,\addrend}) \\
%   \\
%   &H^\nwl : \Addrs^2 \fun \States \fun (\Wor \monstrnefun \UPred{\HeapSegments})\\
%   &H^\nwl_{\start,\addrend} \; s \;\hat{W} \defeq \quad\left\{\npair{\hs} \; \middle| \;
%     \begin{aligned}
%       &n = 0 \text{\footnotesize{ or }} (\dom(\hs) = [\start,\addrend] \text{\footnotesize{ and }}\\
%       &\forall \addr \in [\start,\addrend] \ldotp \\
%       & \;\; {\footnotesize\nonlocal{\ms(\addr)}} \text{\footnotesize{ and }}\\
%       & \;\; \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W})))
%      \end{aligned}
%      \right\}
% \end{align*}
% \caption{Standard regions}
% \label{fig:standard-regions}
% \end{figure}

@@

\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
\dominique{I would try to motivate the LR defs by saying that the LR is
  constructed to express the FTLR and the FTLR says that if a piece of code only
  has access to values respecting the invariants in a world, then it will
  respect them itself too.} The logical relation defines semantically what it
means to be capability safe.\dominique{We have not introduced capability
  safety..} In large parts, capability safety is defined by when a capability
can be safely given to an untrusted piece of code. In the following paragraphs,
we will go through the definitions of the logical relation and provide an
intuition why they define capability safety. The definition of the logical
relation is found in \figurename~\ref{fig:logrel}, and \figurename~\ref{fig:perm-conds}.
Some definitions are omitted and explained when needed. The precise definitions
can be found in the technical appendix~\cite{XXX}.

@@

The \emph{register-file relation} $\stdrr$ specifies that register-files are
safe when they contain safe words (i.e.\ words in $\stdvr$) in all registers but
$\pcreg$. The \emph{expression relation} $\stder$ defines when a word is safe
to use as a program counter. That is the case when the word can be plugged into
a capability safe register file (i.e.\ a register file in $\stdrr$) and paired
with a memory satisfying the world to become a capability safe configuration.

The \emph{value relation} $\stdvr$ defines when words are safe, i.e.\ when they
cannot be used to break the protocols defined by the world. Non-capability data
is always safe because it provides no authority. Capabilities give the authority
to manipulate memory and potentially break memory protocols, so they need to
satisfy certain conditions to be capability safe. In
\figurename~\ref{fig:perm-conds}, we define such a condition for each kind of
permission a capability can have (we will call these permission-based
conditions).

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read capability safe words, by requiring that the region
governing the relevant part of memory, guarantees that the memory contains only
words in the value relation. Generally speaking, global capabilities should only
depend on permanent regions. Therefore we use the function
$\var{localityReg}(\gl,W)$, which projects out all active regions when
$\gl$ (the locality) is local, but only the permanent regions when $\gl$ is global.
The read condition is defined in terms of a standard region
$\iota^\pwl_{\start,\addrend}$ which requires all the words in the range
$[\start,\addrend]$ to be safe.

For a capability with write permission, $\writeCond{}$ must be
satisfied for the capability's range of authority. Such a capability
may be used to write any word, so the region governing the relevant
part of memory must (at least) allow any capability safe word to be
stored. However, this is not entirely true because of the distinction
between write and permit-write-local capabilities.  Only
permit-write-local capabilities need to allow arbitrary safe values to
be stored, while for write capabilities, it is sufficient to allow
storing values that are both non-local and safe. The write condition
uses $\var{localityReg}$ to pick an appropriate region for its
locality in the same way the read condition does. The condition for
the region to be address-stratified is there for technical reasons. In
the value relation, the write condition is used with $\iota^\pwl$ for
write-local capabilities. For write capabilities, the write condition
is used with another standard region $\iota^\nwl_{\start,\addrend}$
which requires all the words in the range $[\start,\addrend]$ to be
non-local and safe.

@@

To prove Lemma~\ref{lem:correctness-f1}, it suffices to show that the start
configuration is capability safe (in the $\observations$ relation) for a world
with a permanent region that requires the assertion flag to be 0. We prove this
using an anti-reduction lemma by which it suffices to show that the
configuration is capability safe after some reduction steps. We then use a general lemma
for reasoning about \texttt{scall}, by which it suffices to show that (1) the
configuration that \texttt{scall} will jump to is capability safe and (2) that
the configuration just after \texttt{scall} is done cleaning up is capability
safe. We use the Fundamental Theorem to reason about the unknown code, but
notice that the Fundamental Theorem says nothing about enter capabilities and
that the adversary capability is an enter capability. Luckily the enter
capability becomes $\exec$ after the jump and then, the Fundamental Theorem
applies.

@@

In \texttt{f3} there are two calls to an adversary and in order for the
assertion in the middle to succeed, they need to be well-bracketed. If the
adversary was able to store the return pointer from the first call and invoke
it in the second call, then \texttt{f3} would have $2$ on top of its stack and
the assertion will fail. However, the security measures of our calling
convention explained in Section~\ref{sec:stack-and-return-pointer} prevent this
sort of attack: specifically, the return pointer is local, so it can only be
stored on the stack, but the part of the stack that is accessible to the
adversary is cleared before the second invocation. The following lemma shows
that there are also no other attacks that will break well-bracketedness of this
example, i.e.\ the assertion never fails. It is similar to the two previous
lemmas:
\begin{figure}[t]
  \centering
  \begin{lstlisting}
f3: push 1
    fetch $r_1$ $\adv$
    scall $r_1$($[]$,$[r_1]$)
    pop $r_2$
    assert $r_2$ 1
    push 2
    scall $r_1$($[]$,$[]$)
    halt
\end{lstlisting}
  \caption{Simple program that relies on control-flow integrity.}
  \label{fig:prog-f3}
\end{figure}

>>>>

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item In Section~\ref{sec:capab-mach-with}, we formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack in Section~\ref{sec:stack-and-return-pointer}. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., (potentially)
  cross-component function pointers) and can be efficient assuming
  only
  one additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed
  logical relation with recursively defined Kripke worlds. It is
  inspired by one used in \cite{Devriese:2016ObjCap} for
  capability safety in a JavaScript-like lambda calculus, but has two
  novel technical characteristics. The natural CPS style of assembly
  code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal
  closure that has been previously used for lambda calculi. Perhaps
  more surprisingly, we show that the special nature of local
  capabilities can be elegantly expressed using a variant of an idea
  that has been used by Dreyer et. al.~\cite{Dreyer:jfp12} to express
  well-bracketedness guarantees in a lambda calculus: Kripke worlds
  with both a public and a private future world relation. Interestingly,
  the two future world relations generally work very similarly to
  Dreyer et al.'s, except that they turn up in quite a different place
  in the logical relation. The logical relation and theorem is presented in Section~\ref{sec:logical-relation}.
\item We demonstrate the power of our results in Section~\ref{sec:examples} by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability safety and our stack management discipline.
\end{itemize}
\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

The $\load{}{}$ instruction loads the contents of the requested memory location
into the appropriate register, but only if the capability has appropriate
authority (i.e. read permission and an appropriate range). The $\restricttwo{}{}$
instruction will update a capability's permissions and global/local status in
the register file, but requires that the new permissions are weaker than the
original and does not turn local capabilities into global ones. The $\geta{}{}$
instruction queries the current address of a capability and stores it in another
register.

@@

The final semantics we discuss is that of $\store{}{}$. As expected, it updates
the memory to the requested value if the capability has write authority for the
requested location. However, the instruction is complicated by the presence of
\emph{local capabilities}, modeled after the ones in
CHERI~\cite{Watson2015Cheri}. Basically, local capabilities are special in that
they can only be kept in registers, i.e. they cannot be stored to memory. This
means that local capabilities can be \emph{temporarily} given to an adversary,
for the duration of an invocation: if we take care to erase the capability from
the register file after control is passed back to us, we can be sure that they
will not have been able to store the capability. However, there is one exception
to the rule above: local capabilities can be stored to a part of memory for
which we have a capability with write-local authority (i.e. permission $\rwl$ or
$\rwlx$). This exception is intended to accomodate regions of memory that behave
like the C stack: stashing away the contents of registers on the stack can be
allowed, even if this includes local capabilities. As long as all capabilities
with write-local authority are themselves local and the stack is cleared after
control is passed back by the adversary, we will see that this does not break
the intended behavior of local capabilities.

@@

\textbf{Program startup}
When our code starts executing, we assume that the language runtime initializes
the memory as follows: a contiguous array of memory is reserved for the stack,
for which we receive a stack pointer in a register, $r_\stk$, that we reserve for this
purpose. We stress that the stack is not built-in, but merely an
abstraction we put on a piece of the memorey. This stack pointer must be local and it must carry $\rwlx$ permission. It is crucial for
our approach that the stack is the only part of memory for which there will ever
be an obtainable capabilities with $\rwlx$ or $\rwl$ capabilities. We expect the language
runtime to enforce this by not handing out such capabilities for other parts of
memory in any situation (i.e. not through malloc, not during loading or linking
etc.). Additionally, in our examples, we typically also assume some static parts
of memory where instructions are stored or static data. Another part of memory
(called the heap) is initially governed by malloc and at program startup, no
other code has capabilities for this memory. Malloc will hand out capabilities
for allocated regions as requested, but it will never hand out capabilities with
write-local capability (i.e. no $\rwlx$ or $\rwl$ permissions). For simplicity,
we assume that memory allocated through malloc cannot be freed.

\textbf{Returning to the adversary} Perhaps the simplest scenario is the case of
returning to the adversary after they invoked our code. In this case, we will
have received a return pointer from them, and we just need to jump to it as
usual. The only thing we need to take care of is to properly clear the registers
before we do (erasing any data we do not want to pass as (part of) the return
value) and similarly. However, it is important to carefully consider what needs
to happen with the stack. Since we may have used the stack for various purposes
(register spilling, storing local state when invoking other functions etc.), we
clearly need to clear that data before returning to the adversary, to prevent
the adversary from gaining access to it.

However, if we are returning from a function that has itself invoked adversary
code, then even the unused part of the stack may still contain data and
capabilities left there by them, including local capabilities since the stack is
store-local. The adversary could take advantage of this by using the unused part
of the stack to store either stack pointers or return pointers from such
previous invocations and then use them when they're no longer supposed to. To
prevent this, before returning, we need to clear (i.e. overwrite with zeros) the
entire part of the stack that the adversary has had access to, not just the
parts that we have used ourselves. Essentially, this is needed for the same
reason that we clear the register file: because the adversary should not be able
to keep hold of local capabilities when control is passed to the trusted code
and back to the adversary. Clearing the entire unused part of the stack is the
most problematic aspect of our discipline in terms of efficiency, since it may
consist of a large part of memory. However, we believe there are ways to
optimize it that we discuss in Section~\ref{sec:discussion}.

@@

A final required safety measure is that before invocation of the adversary, we
need to clear the part of the stack that we are not using. Otherwise, similarly
to what we explained above for returning to the adversary, that part of the
stack may still contain local capabilities from previously executed code and the
adversary could exploit this in much the same way as above.

@@

It's important that neither of these capabilities are accessible to the
adversary. In other words, the return pointer provided to the adversary must be
a capability that they can jump to but not read from, i.e. an enter
capability. To make this work, we construct the activation record as depicted in
Figure~\ref{fig:activ-rec-struct}. The return pointer is an $\entry$ capability
with authority over the activation record containing the previous return and
stack pointer as well as a number of restore instructions. The return pointer's
current address points to those instructions, so that, upon invocation, they are
executed and can load the old stack pointer and program counter back into the
register file. Because the return pointer is given to the adversary as an enter
pointer, they cannot get hold of the contents of the data structure, but after
invocation, its permission is updated to $\exec$, so the activation record's
contents become available to the restore instructions.

The final question that remains is: where should we store this activation
record? The attentive reader may already see that there is only one possibility:
since the activation record contains the old stack pointer, which is necessarily
local, the activation record can only be constructed in a part of memory where
we have write-local access, i.e. on the stack. Note that this means we will be
placing and executing instructions on the stack, i.e. it will not just contain
code pointers and data. This means that our discipline should be combined with
protection against stack smashing attacks (i.e. buffer overflows on the stack
overwriting activation records' contents), but conveniently, the fine-grained
memory protection on capability machines is well-suited to easily provide such
protection.

@@

Our answer is that the adversary should provide it to us, much like we provide
them with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
capability and make sure its current address points just below its start address.
Without this check, an adversary could potentially get control over
our local stack frame during a callback by passing us a local $\rwx$
capability to a global data structure (i.e., there exists a global
capability for it) as the stack pointer and a global
return pointer. If our local state contains no local capabilities,
then following our calling convention, the callback would not fail and
the adversary could use a stored capability for the global data
structure to access our local state. 
To prevent this from happening, we need to make sure the stack
capability is local and there exists no global capability for that
memory segment. By making sure the capability has permission $\rwlx$,
we rely on the system wide assumption that there are only local
$\rwlx$ capabilities.

\textbf{Stack management discipline}
The precautions the needed to safely interact with an adversary have now been introduced and motivated. The stack discipline comprises the precautions. Summarized, they were:
\begin{description}[font=\normalfont\itshape]
\item[At program startup] a local $\rwlx$-capability (the stack pointer) resides in register $r_\stk$. The system does not hand out global write-local capabilities.
\item[Before returning to the adversary] clear the registers (with words we do not want to pass to the adversary) and erase any part of the stack we have used (if any part of the stack has been passed to an adversary, then that part must be erased).
\item[Before invoking the adversary] push an activation record to the stack, create a local $\entry$-capability for the restore instructions of the activation record (the return pointer) and leave it in register $r_0$, leave a copy of the stack capability restricted to the unused part in $r_\stk$, erase the part of the stack we release control over, and clear the non-argument registers.
\item[Before invoking an adversary callback] make sure it is a global capability.
\item[When called by an adversary] make sure the alleged stack pointer has permission $\rwlx$.
\end{description}
\textbf{Reusable macro instructions} 
We define a number of reusable macros. Some of them capture the stack
management discipline and others make the examples in
Section~\ref{sec:examples} readable. The macros are:
\texttt{fetch $r$ $\var{name}$} fetches the capability related to
$\var{name}$ from the linking table and stores it in register
$r$. \texttt{push $r$}/\texttt{pop $r$} works as one would expect. All
macros that uses the stack assumes a stack pointer in register
$r_\stk$.  \texttt{prepstk $r$} is used when a callback is invoked by
the adversary and prepares the received stack pointer by checking that
it has permission $\rwlx$ and making it conform to our stack
discipline. \texttt{scall
  $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)} Jumps
to the capability in register $r$ in the manner described above. That
is, it pushes local state (the contents of registers
$\overline{r_{\var{priv}}}$), stack pointer and return pointer to the
stack, sets up the necessary on stack return code and creates a return
pointer, restricts the stack pointer, clears the unused part of the
stack, clears the necessary registers and jumps to $r$. Upon return,
the private state is restored. \texttt{mclear $r$} clears all the
memory the capability in register $r$ has authority
over. \texttt{rclear $\var{regSet}$} clears all the registers in
$\var{regSet}$. \texttt{reqglob $r$} checks whether the word in
register $r$ is a global capability. \texttt{crtcls
  $\overline{(x_i,r_i)}$ $r$} allocates a closure pair where the
capability in register $r$ is used to point to the program and a new
environment is allocated (using malloc) where the contents of
$\overline{r_i}$ is stored. In the program referred to by $r$, an
implicit fetch happens when an instruction refers to $x_i$.

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
  &\iota_{\start,\addrend}^\pwl : \Regions\\
  &\iota_{\start,\addrend}^\pwl \defeq (\temp,1,=,=,H^\pwl_{\start,\addrend}) \\
  \\
  &H^\pwl : \Addrs^2 \fun \States \fun (\Wor \monwknefun \UPred{\HeapSegments})\\
  &H^\pwl_{\start,\addrend}\; s \; \hat{W} \defeq  \quad\left\{\npair{\hs} \; \middle| \;
    \begin{aligned}
      & n = 0 \text{\footnotesize{ or }} (\dom(\hs) = [\start,\addrend] \text{\footnotesize{ and }} \\
      &\forall \addr \in [\start,\addrend] \ldotp\\
      & \;\; \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W})))
    \end{aligned}
        \right\}\\
  &\iota_{\start,\addrend}^{\nwl} : \Regions \\
  &\iota_{\start,\addrend}^{\nwl} \defeq (\temp,1,=,=,H^\nwl_{start,\addrend}) \\
  \\
  &H^\nwl : \Addrs^2 \fun \States \fun (\Wor \monstrnefun \UPred{\HeapSegments})\\
  &H^\nwl_{\start,\addrend} \; s \;\hat{W} \defeq \quad\left\{\npair{\hs} \; \middle| \;
    \begin{aligned}
      &n = 0 \text{\footnotesize{ or }} (\dom(\hs) = [\start,\addrend] \text{\footnotesize{ and }}\\
      &\forall \addr \in [\start,\addrend] \ldotp \\
      & \;\; {\footnotesize\nonlocal{\ms(\addr)}} \text{\footnotesize{ and }}\\
      & \;\; \npair[n-1]{\hs(\addr)} \in \stdvr(\xi(\hat{W})))
     \end{aligned}
     \right\}
\end{align*}
\caption{Standard regions}
\label{fig:standard-regions}
\end{figure}

@@

\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
\dominique{I would try to motivate the LR defs by saying that the LR is
  constructed to express the FTLR and the FTLR says that if a piece of code only
  has access to values respecting the invariants in a world, then it will
  respect them itself too.} The logical relation defines semantically what it
means to be capability safe.\dominique{We have not introduced capability
  safety..} In large parts, capability safety is defined by when a capability
can be safely given to an untrusted piece of code. In the following paragraphs,
we will go through the definitions of the logical relation and provide an
intuition why they define capability safety. The definition of the logical
relation is found in \figurename~\ref{fig:logrel},
\figurename~\ref{fig:standard-regions}, and \figurename~\ref{fig:perm-conds}.
Some definitions are omitted and explained when needed. The precise definitions
can be found in the technical appendix~\cite{XXX}.

@@

The \emph{register-file relation} $\stdrr$ specifies that register-files are
safe when they contain safe words (i.e., words in $\stdvr$) in all registers but
$\pcreg$. The \emph{expression relation} $\stder$ defines when a word is safe
to use as a program counter. That is the case when the word can be plugged into
a capability safe register file (i.e., a register file in $\stdrr$) and paired
with a memory satisfying the world to become a capability safe configuration.

The \emph{value relation} $\stdvr$ defines when words are safe, i.e. when they
cannot be used to break the protocols defined by the world. Non-capability data
is always safe because it provides no authority. Capabilities give the authority
to manipulate memory and potentially break memory protocols, so they need to
satisfy certain conditions to be capability safe. In
\figurename~\ref{fig:perm-conds}, we define such a condition for each kind of
permission a capability can have (we will call these permission-based
conditions).

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read capability safe words, by requiring that the region
governing the relevant part of memory, guarantees that the memory contains only
words in the value relation. Generally speaking, global capabilities should only
depend on permanent regions. Therefore we use the function
$\var{localityReg}(g,W)$, which projects out all active regions when $g$ is
local, but only the permanent regions when $g$ is global. The read condition
uses $\iota^\pwl_{\start,\addrend}$, a standard regions defined in
Figure~\ref{fig:standard-regions}, which requires all the words in the range
$[\start,\addrend]$ are safe.

For a capability with write permission, $\writeCond{}$ must be satisfied for the
capability's range of authority. Such a capability may be used to write any
word, so the region governing the relevant part of memory must (at least) allow
any capability safe word to be stored. However, this is not entirely true
because of the distinction between write and permit-write-local capabilities.
Only permit-write-local capabilities need to allow arbitrary safe values to be
stored, while for write capabilities, it is sufficient to allow storing values
that are both non-local and safe. The write condition uses $\var{localityReg}$
to pick an appropriate region for its locality in the same way the read
condition does. The condition for the region to be address-stratified is there
for technical reasons.

@@

To prove Lemma~\ref{lem:correctness-f1}, it suffices to show that the start
configuration is capability safe (in the $\observations$ relation) for a world
with a permanent region that requires the assertion flag to be 0. We prove this
using an anti-reduction lemma by which it suffices to show that the
configuration is safe after some reduction steps. We then use a general lemma
for reasoning about \texttt{scall}, by which it suffices to show that (1) the
configuration that \texttt{scall} will jump to is capability safe and (2) that
the configuration just after \texttt{scall} is done cleaning up is capability
safe. We use the Fundamental Theorem to reason about the unknown code, but
notice that the Fundamental Theorem says nothing about enter capabilities and
that the adversary capability is an enter capability. Luckily the enter
capability becomes $\exec$ after the jump and then, the Fundamental Theorem
applies.

@@

In \texttt{f3} there are two calls to an adversary and in order for the
assertion in the middle to succeed, they need to be well-bracketed. If the
adversary were able to store the return pointer from the first call and invoke
it in the second call, then \texttt{f3} would have $2$ on top of its stack and
the assertion will fail. However, the security measures of our calling
convention explained in Section~\ref{sec:stack-and-return-pointer} prevent this
sort of attack: specifically, the return pointer is local, so it can only be
stored on the stack, but the part of the stack that is accessible to the
adversary is cleared before the second invocation. IThe following lemma shows
that there are also no other attacks that will break well-bracketedness of this
example, i.e. the assertion never fails. It is similar to the two previous
lemmas:
\begin{figure}[t]
  \centering
  \begin{lstlisting}
f3: push 1
    fetch $r_1$ $\adv$
    scall $r_1$($[]$,$[r_1]$)
    pop $r_2$
    assert $r_2$ 1
    push 2
    scall $r_1$($[]$,$[]$)
    halt
\end{lstlisting}
  \caption{Simple program that relies on control-flow integrity.}
  \label{fig:prog-f3}
\end{figure}
